The genomic impacts of drift and selection for hybrid performance in maize

### 2015-08-27 response.tex

A third major issue that manuscript seems to be not at all informed by multiple experimental evolution 
studies conducted on Drosophila and other model systems. This contributes to a couple of awkward 
citations of personal communication (p 17 and 19), where ideas about loss of diversity and need for 
replication of experimental populations have been reported in the primary literature. See Drosophila papers 
by Tom Turner and Anthony Long and colleagues.

\response{We have now incorporated mention of additional papers (Drosophila and E. coli) in the context of population size and drift in experimental evolution.  The only remaining personal communication regards the observation of sugary alleles as evidence of contamination, for which there is no publication. }

Finally, the Discussion seems a bit overly long and could potentially be more tightly written.

\response{We have incorporated additional changes requested by reviewers but overall have streamlined the discussion considerably.}

### 2015-08-27 main.tex
accept most jpg changes

\abstract{
Although maize is naturally an outcrossing organism, modern breeding develops of highly inbred lines that are used in controlled crosses to produce hybrids. 
The United States Department of Agriculture reciprocal recurrent selection experiment between the Iowa Stiff Stalk Synthetic (BSSS) and the Iowa Corn Borer Synthetic No. 1 (BSCB1) populations represents one of the longest \rev{running experiments to understand the response to selection} \jpg{changed the wording here, because the experiment's goal was more broad than undesrtanding the genetic basis} for hybrid performance. 
To investigate the genomic impact of this selection program, \rev{we genotyped the progenitor lines and over 600 individuals across multiple cycles of selection using a genome-wide panel of $\sim 40,000$ SNPs.
We confirmed previous results showing a steady temporal decrease in genetic diversity within populations and a corresponding increase in differentiation between populations.  
Thanks to detailed historical information on experimental design, we were able to perform extensive simulations using founder haplotypes to replicate the experiment in the absence of selection.
These simulations demonstrate that while most of the observed reduction in genetic diversity can be attributed to genetic drift,   
heterozygosity in each population has fallen more than expected. 
We then took advantage of our high-density genotype data to identify extensive regions of haplotype fixation and trace haplotype ancestry to single founder inbred lines.
The vast majority of regions showing such evidence of selection differ between the two populations, providing evidence for the dominance model of heterosis.}
We discuss how this pattern is likely to occur during selection for hybrid performance, and how it poses challenges for dissecting the impacts of modern breeding and selection on the maize genome. }

@@

\rev{While the founders of these heterotic groups were not initially differentiated, m}ultiple studies with molecular markers have indicated that these heterotic groups have diverged genetically over time to become highly structured and isolated populations, resulting in a dramatic restructuring of population genetic variation \citep{duvick2004long, ho2005extent, feng2006temporal}. 
Advances in high throughput genotyping and the development of a maize reference genome now enable the observation of maize population structure at high marker density across the whole genome \citep{ganal2011a-large,chia2012maize}. 
These studies have examined a broad spectrum of germplasm at various points in the history of maize to search for the signals of population structure and artificial selection \citep{Hufford2012b, van2012historical,jiao2012genome}. 
Although selective sweeps from the initial domestication of maize are clearly visible, localized genomic signals of selection during modern breeding remain difficult to observe \citep{Hufford2012b,van2012historical} despite steady, heritable improvement in phenotype \citep{duvick2005contribution}. 
The lack of distinct selection signals \rev{in broad germplasm collections may be due to population-specific selection within germplasm subgroups; in this case selection should be easier to detect in populations maintained in an individual program.} 
This \rev{possibliity} is supported by the measured success in identifying targets of selection in individual experimental populations \rev{under directional selection for specific phenotypes such as seed size and ear number} \citep{hirsch2014insights, beissinger2014genome}.

\rev{In this study, we examine a different experiment in which the method and target of selection --- reciprocal recurrent selection for hybrid yield --- closely mirror those used in the generation of modern maize hybrids \citep{comstock1949breeding, duvick2004long}. 
Reciprocal recurrent selection is a method, initially proposed by \citet{comstock1949breeding}, in which lines from two populations are evaluated based on the phenotype of the hybrids each line produces when crossed with the opposing population.
The selected lines are then intermated within each population to generate lines for the next cycle of recurrent selection.
%Reciprocal recurrent selection simultaneously improves two populations for their ability to produce high yielding hybrids when individuals from the two populations are crossed. 
This procedure results in two closed and genetically isolated populations that simultaneously evolve improved combining ability with one another. 
The USDA-ARS at Ames, Iowa has conducted a reciprocal recurrent selection program (hereafter referred to as the Iowa RRS) with the Iowa Stiff Stalk Synthetic (BSSS) and Iowa Corn Borer Synthetic No. 1 (BSCB1) populations for 18 cycles of selection for hybrid yield \citep{penny1971twenty}. \jpg{I think we need Jode's reference here as well if we are going to mention it has gone through 18 cycles.}
This program represents one of the best-documented public experiments on selection for combining ability and hybrid performance. 
Extensive records on population sizes, breeding methods, selection differentials, and seed are available for all cycles of selection. 
This set of resources makes the Iowa RRS an ideal test case for the study of the genomic impact of hybrid breeding in maize.} 
The Iowa RRS experiment \rev{provides additional relevance} because lines derived from the BSSS population have had a major impact on the development of commercial hybrids \citep{duvick2004long, darrah1986}, the formation of modern heterotic groups \citep{troyer1999background, senior1998utility}, and the choice of a maize reference genome \citep{schnable2009the-b73-maize}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO

@@

Our analysis of the Iowa RRS experiment reveals a steady loss of diversity in the BSSS and BSCB1 populations as they became increasingly differentiated from one another over time.
Principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
\rev{Similar observations have been recently made in other reciprocal recurrent selection programs, even in comparisons of multiple replicated populations \citep{romay2012effect,lamkey2014relative}.
These patterns differ markedly, however, from similar experimental evolution in systems such as \emph{Drosophila} \citep{turner2011population, burke2010genome} and \emph{E. coli} \citep{tenaillon2012molecular} in which the effects of selection on diversity are readily discernible. 
A key difference between these studies and those in maize are the effective population sizes, which were kept much lower to produce a short-term phenotypic response to selection within available field testing resources.} \jpg{My intent in rewording this is to emphasize that Ne is the important difference between the scenarios.}

@@

The observation that different targets of selection are observed in opposing heterotic populations bears implications for the genetic mechanisms responsible for heterosis and the success of maize hybrids. 
Classic overdominance models of heterosis predict that at a single locus two distinct alleles confer heterozygote advantage when combined.
Alternatively, the dominance model predicts that heterosis is driven by dominance effects and the complementation of linked alleles in low-recombination regions (dominance or pseudo-overdominance). 
In the case of true overdominance, selection should thus lead to decreased heterozygosity at the same locus in both populations as complementary haplotypes are fixed in each group \rev{\citep[e.g.][]{guo2014maize}}. \jpg{One possible criticism is that overdominance could still be happening when each allele is on multiple founder haplotypes. Could say something like "Although we cannot rule out soft sweeps of complementary overdominant alleles, the observed pattern more parsimoniously favors..."}
We find \rev{little evidence to support this genetic phenomenon, finding only two 2 cM windows genome-wide in which both populations show significantly reduced heterozygosity.} 
The observed pattern instead favors a dominance model, where fixation of a haplotype in one population simply selects against that same haplotype in the other population. 
\rev{Although strongly deleterious variants were likely purged during the inbreeding process leading to the founder lines, many weakly deleterious alleles can be found segregating at low frequencies among inbreds \citep{Mezmouk2014}.}
Because deleterious alleles will be rare in both populations, most haplotypes in the second population will have a different suite of deleterious variants and will complement the fixed haplotype reasonably well.
\rev{We expect that selection against homozygosity of the fixed haplotype will thus} have little impact on diversity in the second population.

@@

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement \citep{duvick2004long}.
\jpg{I'm not sure where this paragraph belongs.}Similar patterns of diversity and population structure can be seen broadly across North American maize germplasm \citep{van2012historical}, suggesting that genetic drift has most likely played a large role in the current genetic structure of modern maize.  
Nonetheless, breeding has contributed substantially to yield improvements \citep{duvick2005contribution}, and the identification and characterization of loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
Several key inbreds in the ‘stiff-stalk’ heterotic group --- B73, B37 and B14 --- were derived from the BSSS population \citep{darrah1986, troyer1999background}. 
%B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tab:fix}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1. 
Our observation of population-specific selection is consistent with a model of selection for complementation in the Iowa RSS experiment, and the lack of extreme differentiation seen over time across US Corn Belt germplasm may suggest a role for such selection in maize hybrid improvement in general \citep{van2012historical}.

>>>>

\abstract{
Although maize is naturally an outcrossing organism, modern breeding utilizes highly inbred lines in controlled crosses to produce hybrids. 
The United States Department of Agriculture reciprocal recurrent selection experiment between the Iowa Stiff Stalk Synthetic (BSSS) and the Iowa Corn Borer Synthetic No. 1 (BSCB1) populations represents one of the longest \rev{running experiments to understand the response to selection} for hybrid performance. 
To investigate the genomic impact of this selection program, \rev{we genotyped the progenitor lines and over 600 individuals across multiple cycles of selection using a genome-wide panel of $\sim 40,000$ SNPs.
We confirmed previous results showing a steady temporal decrease in genetic diversity within populations and a corresponding increase in differentiation between populations.  
Thanks to detailed historical information on experimental design, we were able to perform extensive simulations using founder haplotypes to replicate the experiment in the absence of selection.
These simulations demonstrate that while most of the observed reduction in genetic diversity can be attributed to genetic drift,   
heterozygosity in each population has fallen more than expected. 
We then took advantage of our high-density genotype data to identify extensive regions of haplotype fixation and trace haplotype ancestry to single founder inbred lines.
The vast majority of regions showing such evidence of selection differ between the two populations, providing evidence for the dominance model of heterosis.}
We discuss how this pattern is likely to occur during selection for hybrid performance, and how it poses challenges for dissecting the impacts of modern breeding and selection on the maize genome. }

@@

\rev{While the founders of these heterotic groups were not initially differentiated, m}ultiple studies with molecular markers have indicated that these heterotic groups have diverged genetically over time to become highly structured and isolated populations, resulting in a dramatic restructuring of population genetic variation \citep{duvick2004long, ho2005extent, feng2006temporal}. 
Advances in high throughput genotyping and the development of a maize reference genome now enable the observation of maize population structure at high marker density across the whole genome \citep{ganal2011a-large,chia2012maize}. 
These studies have examined a broad spectrum of germplasm at various points in the history of maize to search for the signals of population structure and artificial selection \citep{Hufford2012b, van2012historical,jiao2012genome}. 
Although selective sweeps from the initial domestication of maize are clearly visible, localized genomic signals of selection during modern breeding are difficult to observe \citep{Hufford2012b,van2012historical} despite steady, heritable improvement in phenotype \citep{duvick2005contribution}. 
The lack of distinct selection signals \rev{in broad germplasm collections may be due to population-specific selection within germplasm subgroups; in this case selection should be easier to detect in populations maintained in an individual program.} 
This possibliity is supported by the measured success in identifying targets of selection in individual experimental populations \rev{under directional selection for specific phenotypes such as seed size and ear number} \citep{hirsch2014insights, beissinger2014genome}.

\rev{In this study, we examine a different experiment in which the method and target of selection --- reciprocal recurrent selection for hybrid yield --- closely mirror those used in the generation of modern maize hybrids \citep{comstock1949breeding, duvick2004long}. 
Reciprocal recurrent selection is a method, initially proposed by \citet{comstock1949breeding}, in which lines from two populations are evaluated based on the phenotype of the hybrids each line produces when crossed with lines from the opposing population.
The selected lines are then intermated within each population to generate lines for the next cycle of recurrent selection.
This procedure results in two closed and genetically isolated populations that simultaneously evolve improved combining ability with one another. 
The USDA-ARS at Ames, Iowa has conducted a reciprocal recurrent selection program (hereafter referred to as the Iowa RRS) with the Iowa Stiff Stalk Synthetic (BSSS) and Iowa Corn Borer Synthetic No. 1 (BSCB1) populations for 18 cycles of selection for hybrid yield \citep{penny1971twenty, edwards2011changes}. 
This program represents one of the best-documented public experiments on selection for combining ability and hybrid performance. 
Extensive records on population sizes, breeding methods, selection differentials, and seed are available for all cycles of selection. 
This set of resources makes the Iowa RRS an ideal test case for the study of the genomic impact of hybrid breeding in maize.} 
The Iowa RRS experiment \rev{provides additional relevance} because lines derived from the BSSS population have had a major impact on the development of commercial hybrids \citep{duvick2004long, darrah1986}, the formation of modern heterotic groups \citep{troyer1999background, senior1998utility}, and the choice of a maize reference genome \citep{schnable2009the-b73-maize}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% INTRO

@@

Our analysis of the Iowa RRS experiment reveals a steady loss of diversity in the BSSS and BSCB1 populations as they became increasingly differentiated from one another over time.
Principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
\rev{Similar observations have been recently made in other reciprocal recurrent selection programs, even in comparisons of multiple replicated populations \citep{romay2012effect,lamkey2014relative}.
These patterns differ markedly, however, from similar experimental evolution in systems such as \emph{Drosophila} \citep{turner2011population, burke2010genome} and \emph{E. coli} \citep{tenaillon2012molecular} in which the effects of selection on diversity are readily discernible. 
A key difference between these studies and those in maize are the effective population sizes, which were kept much lower to produce a short-term phenotypic response to selection within available field testing resources.}

@@

The observation that different targets of selection are observed in opposing heterotic populations bears implications for the genetic mechanisms responsible for heterosis and the success of maize hybrids. 
Classic overdominance models of heterosis predict that at a single locus two distinct alleles confer heterozygote advantage when combined.
Alternatively, the dominance model predicts that heterosis is driven by dominance effects and the complementation of linked alleles in low-recombination regions (dominance or pseudo-overdominance). 
In the case of true overdominance, selection should thus lead to decreased heterozygosity at the same locus in both populations as complementary haplotypes are fixed in each group \rev{\citep[e.g.][]{guo2014maize}}.  
We find \rev{little evidence to support this genetic phenomenon, finding only two 2 cM windows genome-wide in which both populations show significantly reduced heterozygosity. 
Although we cannot rule out soft sweeps of complementary overdominant alleles, the observed pattern more parsimoniously favors} a dominance model, in which fixation of a haplotype in one population simply selects against that same haplotype in the other population. 
\rev{Although strongly deleterious variants were likely purged during the inbreeding process leading to the founder lines, many weakly deleterious alleles can be found segregating at low frequencies among inbreds \citep{Mezmouk2014}.}
Because deleterious alleles will be rare in both populations, most haplotypes in the second population will have a different suite of deleterious variants and will complement the fixed haplotype reasonably well.
\rev{We expect that selection against homozygosity of the fixed haplotype will thus} have little impact on diversity in the second population.

@@

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement \citep{duvick2004long}.
Similar patterns of diversity and population structure can be seen broadly across North American maize germplasm \citep{van2012historical}, suggesting that genetic drift has most likely played a large role in the current genetic structure of modern maize.  
Nonetheless, breeding has contributed substantially to yield improvements \citep{duvick2005contribution}, and the identification and characterization of loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
Several key inbreds in the ‘stiff-stalk’ heterotic group --- B73, B37 and B14 --- were derived from the BSSS population \citep{darrah1986, troyer1999background}. 
%B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tab:fix}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1. 
Our observation of population-specific selection is consistent with a model of selection for complementation in the Iowa RSS experiment, and the lack of extreme differentiation seen over time across US Corn Belt germplasm may suggest a role for such selection in maize hybrid improvement in general \citep{van2012historical}.
